Spatio-angular fluorescence microscopy I. Basic theory (Draft: \today)

### 2018-12-11 papers/basic-theory/paper/paper.tex
Hari and Patrick's edits.

% Talon's custom commands
\setlength\arraycolsep{2.25pt}
\let\originalleft\left
\let\originalright\right
\renewcommand{\left}{\mathopen{}\mathclose\bgroup\originalleft}
\renewcommand{\right}{\aftergroup\egroup\originalright}
\DeclareMathAlphabet{\mathcal}{OMS}{cmsy}{m}{n}
\newcommand{\tensor}[1]{\overset{\text{\tiny$\leftrightarrow$}}{\mb{#1}}}
\newcommand{\argmin}{\arg\!\min}
\newcommand{\me}{\mathrm{e}}
\providecommand{\e}[1]{\ensuremath{\times 10^{#1}}} 
\providecommand{\mb}[1]{\mathbf{#1}}
\providecommand{\msf}[1]{\mathsf{#1}}
\providecommand{\mf}[1]{\mathfrak{#1}}
\providecommand{\mc}[1]{\mathcal{#1}}
\providecommand{\ro}{\mathbf{\mathbf{r}}_o}
\providecommand{\so}{\mathbf{\hat{s}}_o}
\providecommand{\rp}{\mathbf{r}_p}
\providecommand{\rbm}[1]{r_b^{\text{m}}}
\providecommand{\rd}{\mathbf{r}_d}
\providecommand{\rdf}{\mathpzc{r}_d}
\providecommand{\mh}[1]{\mathbf{\hat{#1}}}
\providecommand{\mbb}[1]{\mathbb{#1}}
\providecommand{\bs}[1]{\boldsymbol{#1}}
\providecommand{\bv}{\bs{\nu}}
\providecommand{\taup}{\bs{\tau}}
\providecommand{\bsh}[1]{\hat{\boldsymbol{#1}}}
\providecommand{\nan}{\left(\frac{\text{NA}}{n_o}\right)}
\providecommand{\lmsum}{\sum_{\ell=0}^\infty\sum_{m=-\ell}^{\ell}}
\providecommand{\intr}[1]{\int_{\mbb{R}^{#1}}}
\providecommand{\ints}[1]{\int_{\mbb{S}^{#1}}}
\providecommand{\tb}[1]{\textcolor{blue}{#1}}
% \DeclareFontFamily{OT1}{pzc}{}
% \DeclareFontShape{OT1}{pzc}{m}{it}{<-> s * [1.10] pzcmi7t}{}
% \DeclareMathAlphabet{\mathpzc}{OT1}{pzc}{m}{it}
\newcommand{\eqname}[1]{\tag*{#1}}
\newcommand*\widefbox[1]{\fbox{\hspace{1em}#1\hspace{1em}}}

@@

\author{Talon Chandler,\authormark{1,*} Hari Shroff,\authormark{2,3} Rudolf Oldenbourg,\authormark{2} and Patrick La Rivi\`ere\authormark{1,2}}

@@

\begin{abstract*}
  We introduce the basic elements of a spatio-angular theory of fluorescence
  microscopy. We start by modeling an aplanatic microscope imaging an ensemble
  of in-focus fluorescent dipoles as a linear Hilbert-space operator, and we
  show that the operator takes a particularly convenient form when expressed in
  a basis of spatial and spherical harmonics---a form we call the dipole
  spatio-angular transfer function. We demonstrate our formalism by analyzing a
  single-view fluorescence microscope without using the monopole or scalar
  approximations. We show that this imaging system has a spatio-angular band
  limit, and we exploit the band limit to perform efficient simulations.
  Notably, we show that information about the out-of-plane orientation of
  ensembles of in-focus fluorophores is recorded by paraxial fluorescence
  microscopes.
\end{abstract*}

\section{Introduction}
Fluorescence microscopes are widely used in the biological sciences for
measuring the spatial and angular distribution of ensembles of fluorophores.
While most fluorescence microscopy experiments are designed to measure only the
spatial distribution of fluorophores, a growing number of experiments use
polarizers to measure both the spatial and angular distributions of fluorophores
\cite{vrabioiu2006, mattheyses2010, mehta2016, mcquilken2017, zhanghao2017}.

Meanwhile, single-molecule localization microscopy (SMLM) experiments use
spatially sparse fluorescent samples to localize single molecules with precision
that surpasses the diffraction limit. Real SMLM experiments are limited by noise
\cite{foreman2011, chao2016} and model mismatch, and several studies have shown
that model mismatch (e.g. ignoring the effects of vector optics, dipole
orientation \cite{backlund2014}, and dipole rotation \cite{lew2013}) can
introduce localization bias into SMLM experiments. Therefore, the most precise
SMLM experiments must use an appropriate model and jointly estimate both the
position and orientation of each fluorophore. Several studies have successfully
used vector optics and dipole models to estimate the position and orientation of
single molecules \cite{bohmer2003, lieb2004, toprak2006, aguet2009}, and there
is growing interest in designing optical systems for measuring the position,
orientation, and rotational dynamics of single molecules \cite{agrawal2012,
  backer2014, zhang2018, zhang2018-2}.

While many studies have focused on improving imaging models for spatially sparse
fluorescent samples, we consider the more general case and aim to improve
imaging models for arbitrary samples. In particular, we examine the effects of
two widely used approximations in fluorescence microscopy---the \textit{monopole
  approximation} and the \textit{scalar approximation}.

We use the term \textit{monopole} to refer to a model of a fluorophore that
treats it as an isotropic absorber/emitter. Although the term \textit{monopole
  approximation} is not in widespread use, we think that it is an accurate label
for the way many models of fluorescence microscopes treat fluorophores, and we
use the term to distinguish the monopole model from more realistic dipole and
higher-order models. Despite their use in models, monopole absorber/emitters do
not exist in nature, and the monopole approximation must be justified. Although
many works implicitly apply the monopole approximation, we have encountered two
explicit justifications: (1) the sample contains many randomly oriented
fluorophores within a resolvable volume or (2) the sample contains unconstrained
rotating fluorophores. While both of these situations yield monopole-like
samples, neither yields a sample that is perfectly described by the monopole
model. In this work we investigate the dipole model of fluorophores in detail
and find the conditions under which the monopole approximation is justified.

All fluorescence microscopy techniques that use an \textit{optical point spread
  function} or an \textit{optical transfer function} to model the mapping
between the fluorophores and the irradiance on the detector implicitly make the
monopole approximation. The optical point spread function is the irradiance
response of an optical system to an isotropic point source, so it cannot model
the response due to an anisotropic dipole radiator. In this work we define
\textit{monopole and dipole transfer functions} that describe the mapping
between fluorophores and the measured irradiance. Although optical systems are
an essential part of microscopes, fluorescence microscopists are interested in
measuring fluorophores (not optics), so the monopole and dipole transfer
functions are more valuable than the optical transfer function.

While the monopole approximation applies to the fluorescent object, the
\textit{scalar approximation} applies to the fields that propagate through the
microscope. Modeling the electric fields in a region requires a
three-dimensional vector field, but if the electric fields are random or
parallel a scalar field is sufficient. Replacing the vector-valued electric
field, $\mb{E}$, with a scalar-valued field, $U$, is the scalar approximation,
and this approximation is only justified if the fields are randomly or uniformly
polarized.

The scalar approximation is often made together with the monopole approximation.
For example, the Born-Wolf model \cite{born1980} and the Gibson-Lanni model
\cite{gibson89} make both the monopole and scalar approximations when applied to
fluorescence microscopes. However, some models make the monopole approximation
but not the scalar approximation. For example, the Richards-Wolf model
\cite{richards} considers the role of vector-valued fields in the optical
system, but it is an optical model so when it is applied to fluorescence
microscopes the monopole approximation is required. As previously mentioned,
several works in the SMLM community have modeled and applied dipole point spread
functions that do not make the monopole or scalar approximations, but to our
knowledge no works have considered dipole transfer functions.

This work lies at the intersection of three fluorescence microscopy communities:
(1) the spatial ensemble imaging community where many fluorophores are imaged in
parallel and the goal is to find their concentration throughout the sample, (2)
the spatio-angular ensemble community where many fluorophores are imaged in
parallel and the goal is to find their concentration and orientation, and (3)
the single molecule localization microscopy (SMLM) community where fluorophores
are sparse in the sample and the goal is to find the position of each
fluorophore. We briefly review how these three communities use the monopole and
scalar approximations.

@@

A small but growing group of microscopists is interested in measuring the
orientation and position of ensembles of fluorophores \cite{vrabioiu2006,
  mattheyses2010, mehta2016, mcquilken2017, zhanghao2017}. These techniques
typically use polarizers to make multiple measurements of the same object, then
they use a model of the dipole excitation and emission processes
\cite{fourkas2001} to recover the orientation of fluorophores using pixel-wise
arithmetic. Although these studies do not adopt the scalar or monopole
approximations for the angular part of the problem, they adopt both
approximations when they consider the spatial part of the problem. Existing
works either ignore the spatial reconstruction problem \cite{vrabioiu2006,
  mattheyses2010, mehta2016, mcquilken2017} or assume that the spatial and
angular reconstruction problems can be performed sequentially
\cite{zhanghao2017}.

@@

\begin{figure}
  \centering
  \includegraphics[scale=1.0]{../figures/monopole-block/monopole-block.pdf}
  \caption{The mapping between the object and data space of a monopole
    fluorescence microscope can be computed in two different bases---a delta
    function basis and a spatial harmonic basis. The change of basis can be
    computed with a two-dimensional Fourier transform denoted
    $\mathcal{F}_{\mbb{R}^2}$.}
     \label{fig:monopole-block}      
\end{figure}

The mapping between the object and the data in a linear shift-invariant imaging
system takes a particularly simple form in a spatial harmonic (i.e. Fourier) basis. If we apply
the Fourier convolution theorem to Eq. \eqref{eq:lsi} we find that
\begin{align}
  G(\bv) = H(\bv)F(\bv),\label{eq:freq}
\end{align}
where we define the \textit{scaled irradiance spectrum} as
\begin{align}
  G(\bv) = \int_{\mbb{R}^2}d\mb{r}\, g(\mb{r})\, \text{exp}(-2\pi i\mb{r}\cdot\bv),
\end{align}
the \textit{monopole transfer function} as
\begin{align}
  H(\bv) = \int_{\mbb{R}^2}d\mb{r}\, h(\mb{r})\, \text{exp}(-2\pi i\mb{r}\cdot\bv),\label{eq:otf}
\end{align}
and the \textit{monopole spectrum} as
\begin{align}
    F(\bv) = \int_{\mbb{R}^2}d\mb{r}\, f(\mb{r})\, \text{exp}(-2\pi i\mb{r}\cdot\bv).
\end{align}
The monopole point spread function is normalized and real, so we know that the
monopole transfer function is normalized, $H(0) = 1$, and conjugate symmetric,
$H(-\bv) = H^*(\bv)$, where $z^*$ denotes the complex conjugate of $z$.

@@

\subsubsection{Monopole coherent transfer functions}
Although the Fourier transform can be used to calculate the monopole transfer
function directly from the monopole point spread function, there is a well-known
alternative that exploits coherent transfer functions. The key idea is that the
monopole point spread function can always be written as the absolute square of a
scalar-valued \textit{monopole coherent spread function} defined as
\begin{align}
  |c(\rd - \ro)|^2 = h(\rd - \ro). \label{eq:absquarescalar}
\end{align}
Physically, the monopole coherent spread function corresponds to the
scalar-valued field on the detector with appropriate scaling.

We can plug Eq. \eqref{eq:absquarescalar} into Eq. \eqref{eq:otf} and use the
autocorrelation theorem to rewrite the monopole transfer function as
\begin{align}
  H(\bv) = \int_{\mbb{R}^2}d\taup\, C(\taup)C^*(\taup - \bv), 
\end{align}
where we have introduced the
\textit{monopole coherent transfer function} as the two-dimensional Fourier
transform of the monopole coherent spread function
\begin{align}
  C(\taup) = \int_{\mbb{R}^2}d\mb{r}\, c(\mb{r})\,\text{exp}[-2\pi i\mb{r}\cdot\taup].
\end{align}
Physically, the monopole coherent transfer function corresponds to the
scalar-valued field in a Fourier plane of the detector with appropriate scaling.
% We have chosen the symbol $\taup$ as the coordinate for the coherent
% transfer function because the Fourier plane of the detector is often called the
% \textit{pupil plane}.

@@

In a delta function basis the object can be represented by a function
$f(\ro, \so)$ called the \textit{dipole density}---the number of dipoles per
unit area and per unit solid angle at position $\ro{}$ and oriented along
$\so{}$. Similar to the monopole case, we model the mapping between the object
and the irradiance in a delta function basis as an integral transform
\begin{align}
  g'(\rd') = \int_{\mbb{S}^2}d\so\int_{\mbb{R}^2}d\ro\, h'(\rd', \ro, \so)f(\ro, \so),
\end{align}
where $h'(\rd', \ro, \so)$ is the irradiance at position $\rd'$ created by a
point source at $\ro$ with orientation $\so$. If the optical system is
aplanatic, we can write the integral transform as
\begin{align}
  g'(\rd') = \int_{\mbb{S}^2}d\so\int_{\mbb{R}^2}d\ro\, h'(\rd' - m\ro, \so)f(\ro, \so). 
\end{align}

We define the same demagnified detector coordinate $\rd = \rd'/m$ and a new
normalization factor that corresponds to the total power incident on the
detector due to a spatial point source with an angularly uniform distribution of
dipoles
$P_\text{dip} = \int_{\mbb{S}^2}d\mh{s}\int_{\mbb{R}^2}d\mb{r}\, h'(m\mb{r},
\so)$. We use these scaling factors to define the
\textit{dipole point spread function} as
\begin{align}
  h(\rd - \ro, \so) = \frac{h'(m[\rd - \ro], \so)}{P_\text{dip}}, 
\end{align}
and the \textit{scaled irradiance} as 
\begin{align}
  g(\rd) = \frac{g'(m\rd)}{P_\text{dip}}. 
\end{align}
With these definitions we can express the mapping between the object and the
data as
\begin{align}
g(\rd{}) = \int_{\mbb{S}^2}d\so{}\int_{\mbb{R}^2}d\ro{}\, h(\rd{} -\ro{}, \so{})f(\ro, \so). \label{eq:odpsf}
\end{align}
Equation \eqref{eq:odpsf} is a key result because it represents the mapping
between object space and data space in a delta function basis. The integrals in
Eq. \eqref{eq:odpsf} would be extremely expensive to compute for an arbitrary
object, but the integrals simplify to an efficient sum if the object is
spatially and angularly sparse. In other words, Eq. \eqref{eq:odpsf} is ideal for
simulating and analyzing single fluorophores that are rigidly attached to an
oriented structure.

@@

  \begin{figure}
  \hspace{-5em}
  \includegraphics[scale=1.0]{../figures/dipole-block/dipole-block.pdf}
  \caption{The mapping between the object and data space of a dipole imaging
    system can be computed in four different bases---a delta function basis, a
    spatial-harmonic/angular-delta basis, a spatial-delta/spherical-harmonic
    basis, and a spatio-angular harmonic basis. The changes of basis can be
    computed with the two-dimensional Fourier transform denoted
    $\mathcal{F}_{\mbb{R}^2}$, and the spherical Fourier transform denoted
    $\mathcal{F}_{\mbb{S}^2}$.}
   \label{fig:dipole-block}      
    \end{figure}

@@

  This basis is well suited for simulating and analyzing arbitrary samples
  because it exploits the band limit of the imaging system. We will see the
  advantage explicitly in Section \ref{sec:results} when we calculate a specific
  dipole spatio-angular transfer function and see that it has spatial and
  angular band limits. We note that most single molecules imaging experiments
  are best described in this basis because of the effects of spatial and
  rotational diffusion.

@@

\subsubsection{Dipole coherent transfer functions}
Similar to the monopole case, there is efficient way to calculate the transfer
functions using coherent transfer functions. The dipole point spread function
can always be written as the absolute square of a vector-valued function called
the \textit{dipole coherent spread function}
\begin{align}
  |\mb{c}(\rd - \ro, \so)|^2 = h(\rd - \ro, \so). \label{eq:absquare2}
\end{align}
Physically, the dipole coherent spread function corresponds to the vector-valued
electric field on the detector with appropriate scaling. We need a vector-valued
coherent transfer function since the polarization of the field plays a
significant role in dipole imaging, so the dipole point spread function cannot
be written as an absolute square of a scalar-valued function.

@@

We can plug Eq. \eqref{eq:absquare2} into Eq. \eqref{eq:dstf} and use the
autocorrelation theorem to rewrite the dipole spatial transfer function as
\begin{align}
  H(\bv, \so) = \int_{\mbb{R}^2}d\taup\,\mb{C}(\taup, \so)\mb{C}^\dagger(\taup - \bv, \so), 
\end{align}
where we have introduced the \textit{dipole coherent transfer function}
$\mb{C}(\taup, \so)$ as the two-dimensional Fourier transform of the dipole
coherent spread function
\begin{align}
  \mb{C}(\taup, \so) = \int_{\mbb{R}^2}d\mb{r}\, \mb{c}(\mb{r}, \so)\,\text{exp}[-2\pi i\mb{r}\cdot\taup].
\end{align}
Physically, the dipole coherent transfer function corresponds to the
vector-valued electric field created by a dipole oriented along $\so$ in a
Fourier plane of the detector with appropriate scaling. Similar to the monopole
case, we can calculate the dipole-orientation-dependent fields in a Fourier plane of
the detector, scale appropriately to find the dipole coherent transfer function,
then use the relationships in Fig. \ref{fig:transfer-functions} to calculate
the other transfer functions.

>>>>

% Talon's custom commands
\setlength\arraycolsep{1.5pt}
\let\originalleft\left
\let\originalright\right
\renewcommand{\left}{\mathopen{}\mathclose\bgroup\originalleft}
\renewcommand{\right}{\aftergroup\egroup\originalright}
\DeclareMathAlphabet{\mathcal}{OMS}{cmsy}{m}{n}
\newcommand{\tensor}[1]{\overset{\text{\tiny$\leftrightarrow$}}{\mb{#1}}}
\newcommand{\argmin}{\arg\!\min}
\newcommand{\me}{\mathrm{e}}
\providecommand{\e}[1]{\ensuremath{\times 10^{#1}}} 
\providecommand{\mb}[1]{\mathbf{#1}}
\providecommand{\msf}[1]{\mathsf{#1}}
\providecommand{\mf}[1]{\mathfrak{#1}}
\providecommand{\mc}[1]{\mathcal{#1}}
\providecommand{\ro}{\mathbf{\mathbf{r}}_o}
\providecommand{\so}{\mathbf{\hat{s}}_o}
\providecommand{\rp}{\mathbf{r}_p}
\providecommand{\rbm}[1]{r_b^{\text{m}}}
\providecommand{\rd}{\mathbf{r}_d}
\providecommand{\rdf}{\mathpzc{r}_d}
\providecommand{\mh}[1]{\mathbf{\hat{#1}}}
\providecommand{\mbb}[1]{\mathbb{#1}}
\providecommand{\bs}[1]{\boldsymbol{#1}}
\providecommand{\bv}{\bs{\nu}}
\providecommand{\taup}{\bs{\tau}}
\providecommand{\bsh}[1]{\hat{\boldsymbol{#1}}}
\providecommand{\nan}{\left(\frac{\text{NA}}{n_o}\right)}
\providecommand{\lmsum}{\sum_{\ell=0}^\infty\sum_{m=-\ell}^{\ell}}
\providecommand{\intr}[1]{\int_{\mbb{R}^{#1}}}
\providecommand{\ints}[1]{\int_{\mbb{S}^{#1}}}
\providecommand{\tb}[1]{\textcolor{blue}{#1}}
% \DeclareFontFamily{OT1}{pzc}{}
% \DeclareFontShape{OT1}{pzc}{m}{it}{<-> s * [1.10] pzcmi7t}{}
% \DeclareMathAlphabet{\mathpzc}{OT1}{pzc}{m}{it}
\newcommand{\eqname}[1]{\tag*{#1}}
\newcommand*\widefbox[1]{\fbox{\hspace{1em}#1\hspace{1em}}}

@@

\author{Talon Chandler,\authormark{1,*} Hari Shroff,\authormark{2,3} Rudolf Oldenbourg,\authormark{3} and Patrick La Rivi\`ere\authormark{1,3}}

@@

\begin{abstract*}
  We introduce the basic elements of a spatio-angular theory of fluorescence
  microscopy. We start by modeling an aplanatic microscope imaging an ensemble
  of in-focus fluorescent dipoles as a linear Hilbert-space operator, and we
  show that the operator takes a particularly convenient form when expressed in
  a basis of complex exponentials and spherical harmonics---a form we call the
  dipole spatio-angular transfer function. We demonstrate our formalism by
  analyzing a single-view fluorescence microscope without using the monopole or
  scalar approximations. We show that this imaging system has a spatio-angular
  band limit, and we exploit the band limit to perform efficient simulations.
  Notably, we show that information about the out-of-plane orientation of
  ensembles of in-focus fluorophores is recorded by paraxial fluorescence
  microscopes.
\end{abstract*}

\section{Introduction}
Fluorescence microscopes are widely used in the biological sciences for
measuring ensembles of fluorescent molecules. While most fluorescence microscopy
experiments are designed to measure only the spatial distribution of
fluorophores, a growing number of experiments seek to measure both the spatial
and angular distributions of fluorophores by the use of polarizers
\cite{vrabioiu2006, mattheyses2010, mehta2016, mcquilken2017, zhanghao2017} or
point spread function engineering \cite{agrawal2012, zhang2018}.

Meanwhile, single-molecule localization microscopy (SMLM) experiments use
spatially sparse fluorescent samples to localize single molecules with precision
that surpasses the diffraction limit. Noise limits the precision of this
localization \cite{foreman2011, chao2016}, and several studies have shown that
model mismatch (e.g. ignoring the effects of vector optics, dipole orientation
\cite{backlund2014}, and dipole rotation \cite{lew2013}) can introduce
localization bias as well. Therefore, the most precise and accurate SMLM
experiments must use an appropriate model and jointly estimate both the position
and orientation of each fluorophore. Several studies have successfully used
vector optics and dipole models to estimate the position and orientation of
single molecules \cite{bohmer2003, lieb2004, toprak2006, aguet2009}, and there
is growing interest in designing optical systems for measuring the position,
orientation, and rotational dynamics of single molecules \cite{agrawal2012,
  backer2014, stallinga2015, zhang2018, zhang2018-2}.

While many studies have focused on improving imaging models for spatially sparse
fluorescent samples, we consider the more general case and aim to improve
imaging models for arbitrary samples including those containing ensembles of
fluorophores within a resolvable volume. In particular, we examine the effects
of two widely used approximations in fluorescence microscopy---the
\textit{monopole approximation} and the \textit{scalar approximation}.

We use the term \textit{monopole} to refer to a model of a fluorophore that
treats it as an isotropic absorber/emitter. Although the term \textit{monopole
  approximation} is not in widespread use, we think it accurately describes the
way many models of fluorescence microscopes treat fluorophores, and we use the
term to distinguish the monopole model from more realistic dipole and
higher-order models. Despite their use in models, monopole absorber/emitters do
not exist in nature. All physical fluorophores absorb and emit radiation with
dipole or higher-order transitions, and these transitions are always oriented in
space. For a classical mental model of fluorophores we imagine each dipole as a
small oriented antenna (with incoherent absorption and emission moments) where
electrons are constrained to move along a single direction. With this model we
can argue classically that monopole radiators do not exist because they would
violate the conservation of charge.

Although many works implicitly apply the monopole approximation, we have
encountered two explicit justifications: (1) the sample contains many randomly
oriented fluorophores within a resolvable volume or (2) the sample contains
unconstrained rotating fluorophores. While both of these situations yield
monopole-like samples, neither yields a sample that is perfectly described by
the monopole model. In this work we investigate the dipole model of fluorophores
in detail and find the conditions under which the monopole approximation is
justified.

All fluorescence microscopy techniques that use an \textit{optical point spread
  function} or an \textit{optical transfer function} to model the mapping
between the fluorophores and the irradiance on the detector implicitly make the
monopole approximation. The optical point spread function is the irradiance
response of an optical system to an isotropic point source, so it cannot model
the response due to an anisotropic dipole radiator. In this work we define
\textit{monopole and dipole transfer functions} that describe the mapping
between fluorophores and the measured irradiance. Although optical systems are
an essential part of microscopes, fluorescence microscopists are interested in
measuring the properties of fluorophores (not optics), so the monopole and
dipole transfer functions are more directly useful than the optical transfer
function for the problems that fluorescence microscopists are interested in
solving.

While the monopole approximation applies to the fluorescent object, the
\textit{scalar approximation} applies to the fields that propagate through the
microscope. Modeling the electric fields in a region requires a
three-dimensional vector field, but if the electric fields are random or
parallel a scalar field is sufficient, and we can replace the vector-valued electric
field, $\mb{E}$, with a scalar-valued field, $U$.%  is the scalar approximation,
% and this approximation is only justified if the fields are randomly or uniformly
% polarized.

The scalar approximation is often made together with the monopole approximation.
For example, the Born-Wolf model \cite{born1980} and the Gibson-Lanni model
\cite{gibson89} make both the monopole and scalar approximations when applied to
fluorescence microscopes. However, some models make the monopole approximation
but not the scalar approximation. For example, the Richards-Wolf model
\cite{richards} considers the role of vector-valued fields in the optical
system, but it is an optical model so when it is applied to fluorescence
microscopes the monopole approximation is assumed. % As previously mentioned,
% several works in the SMLM community have modeled and applied dipole point spread
% functions that do not make the monopole or scalar approximations, but to our
% knowledge no works have considered dipole transfer functions.

This work lies at the intersection of three fluorescence microscopy communities:
(1) the spatial ensemble imaging community where many fluorophores are imaged in
parallel and the goal is to find their concentration throughout the sample, (2)
the spatio-angular ensemble community where many fluorophores are imaged in
parallel and the goal is to find their concentration and orientation, and (3)
the SMLM community where fluorophores are sparse in the sample and the goal is
to find the position of each fluorophore. We briefly review how these three
communities use the monopole and scalar approximations.

@@

A small but growing group of microscopists is interested in measuring the
orientation and position of ensembles of fluorophores \cite{vrabioiu2006,
  mattheyses2010, mehta2016, mcquilken2017, zhanghao2017}. These techniques
typically use polarizers to make multiple measurements of the same object, then
they use a model of the dipole excitation and emission processes
\cite{fourkas2001} to recover the orientation of fluorophores using pixel-wise
arithmetic. Although these studies do not adopt the scalar or monopole
approximations for the angular part of the problem, they adopt both
approximations when they consider the spatial part of the problem. Existing
works either ignore the spatial reconstruction problem \cite{vrabioiu2006,
  mattheyses2010, mehta2016, mcquilken2017} or assume that the spatial and
angular reconstruction problems can be solved sequentially
\cite{zhanghao2017}.

@@

\begin{figure}
  \centering
  \includegraphics[scale=1.0]{../figures/monopole-block/monopole-block.pdf}
  \caption{The mapping between the object and data space of a monopole
    fluorescence microscope can be computed in two different bases---a delta
    function basis and a complex exponential basis. The change of basis can be
    computed with a two-dimensional Fourier transform denoted
    $\mathcal{F}_{\mbb{R}^2}$.}
     \label{fig:monopole-block}      
\end{figure}

The mapping between the object and the data in a linear shift-invariant imaging
system takes a particularly simple form in a complex exponential (i.e. Fourier)
basis. If we apply the Fourier convolution theorem to Eq. \eqref{eq:lsi} we find
that
\begin{align}
  G(\bv) = H(\bv)F(\bv),\label{eq:freq}
\end{align}
where we define the \textit{scaled irradiance spectrum} as
\begin{align}
  G(\bv) = \int_{\mbb{R}^2}d\mb{r}\, g(\mb{r})\, \text{exp}(-2\pi i\mb{r}\cdot\bv),
\end{align}
the \textit{monopole transfer function} as
\begin{align}
  H(\bv) = \int_{\mbb{R}^2}d\mb{r}\, h(\mb{r})\, \text{exp}(-2\pi i\mb{r}\cdot\bv),\label{eq:otf}
\end{align}
and the \textit{monopole spectrum} as
\begin{align}
    F(\bv) = \int_{\mbb{R}^2}d\mb{r}\, f(\mb{r})\, \text{exp}(-2\pi i\mb{r}\cdot\bv).
\end{align}
The monopole point spread function is normalized and real, so we know that the
monopole transfer function is normalized, $H(0) = 1$, and conjugate symmetric,
$H(-\bv) = H^*(\bv)$, where $z^*$ denotes the complex conjugate of $z$.

@@

\subsubsection{Monopole coherent transfer functions}
Although the Fourier transform can be used to calculate the monopole transfer
function directly from the monopole point spread function, there is a well-known
alternative that exploits coherent transfer functions. The key idea is that the
monopole point spread function can always be written as the absolute square of a
scalar-valued \textit{monopole coherent spread function}, $c(\rd - \ro)$,
defined by
\begin{align}
  |c(\rd - \ro)|^2 = h(\rd - \ro). \label{eq:absquarescalar}
\end{align}
Physically, the monopole coherent spread function corresponds to the
scalar-valued field on the detector with appropriate scaling.

We can plug Eq. \eqref{eq:absquarescalar} into Eq. \eqref{eq:otf} and use the
autocorrelation theorem to rewrite the monopole transfer function as
\begin{align}
  H(\bv) = \int_{\mbb{R}^2}d\taup\, C(\taup)C^*(\taup - \bv), 
\end{align}
where we have introduced the
\textit{monopole coherent transfer function} as the two-dimensional Fourier
transform of the monopole coherent spread function:
\begin{align}
  C(\taup) = \int_{\mbb{R}^2}d\mb{r}\, c(\mb{r})\,\text{exp}[-2\pi i\mb{r}\cdot\taup].
\end{align}
Physically, the monopole coherent transfer function corresponds to the
scalar-valued field in a Fourier plane of the detector with appropriate scaling.
% We have chosen the symbol $\taup$ as the coordinate for the coherent
% transfer function because the Fourier plane of the detector is often called the
% \textit{pupil plane}.

@@

In a delta function basis the object can be represented by a function
$f(\ro, \so)$ called the \textit{dipole density}---the number of dipoles per
unit area and per unit solid angle at position $\ro{}$ and oriented along
$\so{}$. Similar to the monopole case, we model the mapping between the object
and the irradiance in a delta function basis as an integral transform
\begin{align}
  g'(\rd') = \int_{\mbb{S}^2}d\so\int_{\mbb{R}^2}d\ro\, h'(\rd', \ro, \so)f(\ro, \so),
\end{align}
where $h'(\rd', \ro, \so)$ is the irradiance at position $\rd'$ created by a
point source at $\ro$ with orientation $\so$. Notice that we have considered all
possible orientations $\so$ and integrated over the sphere $\mbb{S}^2$. The
dipole density is always symmetric under angular inversion,
$f(\ro, \so) = f(\ro,-\so)$, so we could have chosen to integrate over a
hemisphere and adjusted the definition the dipole density by a factor of two.
For convenience we will continue to integrate over the complete sphere. We note
that all functions in this work with $\so$ as an independent variable are
symmetric under angular inversion, $\so \rightarrow -\so$.

If the optical system is aplanatic, we can write the
integral transform as
\begin{align}
  g'(\rd') = \int_{\mbb{S}^2}d\so\int_{\mbb{R}^2}d\ro\, h'(\rd' - m\ro, \so)f(\ro, \so). 
\end{align}

We define the same demagnified detector coordinate $\rd = \rd'/m$ and a new
normalization factor that corresponds to the total power incident on the
detector due to a spatial point source with an angularly uniform distribution of
dipoles
$P_\text{dip} = \int_{\mbb{S}^2}d\so\int_{\mbb{R}^2}d\mb{r}\, h'(m\mb{r},
\so)$. We use these scaling factors to define the
\textit{dipole point spread function} as
\begin{align}
  h(\rd - \ro, \so) = \frac{h'(m[\rd - \ro], \so)}{P_\text{dip}}, 
\end{align}
and the \textit{scaled irradiance} as 
\begin{align}
  g(\rd) = \frac{g'(m\rd)}{P_\text{dip}}. 
\end{align}
With these definitions we can express the mapping between the object and the
data as
\begin{align}
g(\rd{}) = \int_{\mbb{S}^2}d\so{}\int_{\mbb{R}^2}d\ro{}\, h(\rd{} -\ro{}, \so{})f(\ro, \so). \label{eq:odpsf}
\end{align}
Equation \eqref{eq:odpsf} is a key result because it represents the mapping
between object space and data space in a delta function basis. The integrals in
Eq. \eqref{eq:odpsf} would be extremely expensive to compute for an arbitrary
object, but the integrals simplify to an efficient sum if the object is
spatially and angularly sparse. In other words, Eq. \eqref{eq:odpsf} is ideal for
simulating and analyzing single fluorophores that are rigidly attached to an
oriented structure.

@@

  \begin{figure}
  \hspace{-5em}
  \includegraphics[scale=1.0]{../figures/dipole-block/dipole-block.pdf}
  \caption{The mapping between the object space and data space of a dipole
    imaging system can be computed in four different bases---a delta function
    basis, a complex-exponential/angular-delta basis, a
    spatial-delta/spherical-harmonic basis, and a
    complex-exponential/spherical-harmonic basis. The changes of basis can be
    computed with the two-dimensional Fourier transform denoted
    $\mathcal{F}_{\mbb{R}^2}$, and the spherical Fourier transform denoted
    $\mathcal{F}_{\mbb{S}^2}$.}
   \label{fig:dipole-block}      
    \end{figure}

@@

  This basis is well suited for simulating and analyzing arbitrary samples
  because it exploits the band limit of the imaging system. We will see the
  advantage explicitly in Section \ref{sec:results} when we calculate a specific
  dipole spatio-angular transfer function and see that it has spatial and
  angular band limits. We note that most single molecule imaging experiments
  are best described in this basis because of the effects of spatial and
  rotational diffusion.

@@

\subsubsection{Dipole coherent transfer functions}
Similar to the monopole case, there is an efficient way to calculate the
transfer functions using coherent transfer functions. The dipole point spread
function can always be written as the absolute square of a vector-valued
function, $\mb{c}(\rd - \ro, \so)$, called the \textit{dipole coherent spread
  function}:
\begin{align}
  |\mb{c}(\rd - \ro, \so)|^2 = h(\rd - \ro, \so). \label{eq:absquare2}
\end{align}
Physically, the dipole coherent spread function corresponds to the vector-valued
electric field on the detector with appropriate scaling. We need a vector-valued
coherent transfer function since the polarization of the field plays a
significant role in dipole imaging, so the dipole point spread function cannot
be written as an absolute square of a scalar-valued function.

@@

We can plug Eq. \eqref{eq:absquare2} into Eq. \eqref{eq:dstf} and use the
autocorrelation theorem to rewrite the dipole spatial transfer function as
\begin{align}
  H(\bv, \so) = \int_{\mbb{R}^2}d\taup\,\mb{C}(\taup, \so)\mb{C}^\dagger(\taup - \bv, \so), 
\end{align}
where we have introduced the \textit{dipole coherent transfer function}
$\mb{C}(\taup, \so)$ as the two-dimensional Fourier transform of the dipole
coherent spread function:
\begin{align}
  \mb{C}(\taup, \so) = \int_{\mbb{R}^2}d\mb{r}\, \mb{c}(\mb{r}, \so)\,\text{exp}[-2\pi i\mb{r}\cdot\taup].
\end{align}
Physically, the dipole coherent transfer function corresponds to the
vector-valued electric field created by a dipole oriented along $\so$ in a
Fourier plane of the detector with appropriate scaling. Similar to the monopole
case, we can calculate the dipole-orientation-dependent fields in a Fourier
plane of the detector, scale appropriately to find the dipole coherent transfer
function, then use the relationships in Fig. \ref{fig:transfer-functions} to
calculate the other transfer functions. Finally, we note that the dipole
coherent transfer function is identical to what Agrawal et. al. call the
\textit{Green's tensor} \cite{agrawal2012} and Novotny and Hecht's
\textit{dyadic point spread function} multiplied by the dipole moment vector
\cite{nov2006}.
